---
title: "Transit Oriented Development in the District of Columbia"
author: "Johnathan Clementi"
date: '9/24/2021'
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
Purpose: Prepare a policy brief for local City Council representatives. Do households value transit-rich neighborhoods compared to others? How certain can you be about your conclusions given some of the spatial biases we’ve discussed?

Details: Provide a brief motivation at the beginning, annotate each visualization appropriately, and then provide brief policy-relevant conclusions. Please show all code blocks. Here are the specific deliverables:

1. Show your data wrangling work.
2. Four small-multiple (2000 & 2017+) visualizations comparing four selected Census variables across time and space (TOD vs. non-TOD).
3. One grouped bar plot making these same comparisons.
4. One table making these same comparisons.
5. Create two graduated symbol maps of population and rent within 0.5 mile of each transit station. Google for more information, but a graduate symbol map represents quantities for each transit station proportionally.
6. Create a geom_line plot that shows mean rent as a function of distance to subway stations (Figure 1.17). To do this you will need to use the multipleRingBuffer function found in the functions.R script.
7. Download and wrangle point-level crime data (pick a crime type). What is the relationship between crime, transit access and rents?


# Setup

#### Load Libraries
```{r setup_package, warning = FALSE, message = FALSE}
rm(list=ls()) ##Clear my objects from memory

library(tidyverse)
library(tidycensus)
library(sf)
library(tmap) # mapping, install if you don't have it
library(kableExtra)
library(rgdal)

options(scipen=999)
options(tigris_class = "sf")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac") #colorwheel
```

#### Load Styling Options
```{r}
mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 16,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.text.x = element_text(size = 14))
}

plotTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 16,colour = "black"),
    plot.subtitle = element_text(face="italic"),
    plot.caption = element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey80", size = 0.1),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.background = element_rect(fill = "grey80", color = "white"),
    strip.text = element_text(size=12),
    axis.title = element_text(size=12),
    axis.text = element_text(size=10),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"),
    strip.text.x = element_text(size = 14)
  )
}
```
#### Load distribution break functions
```{r}
qBr <- function(df, variable, rnd) {
  if (missing(rnd)) {
    as.character(quantile(round(df[[variable]],0),
                          c(.01,.2,.4,.6,.8), na.rm=T))
  } else if (rnd == FALSE | rnd == F) {
    as.character(formatC(quantile(df[[variable]],
                 c(.01,.2,.4,.6,.8), na.rm=T), digits = 3))
  }
}

q5 <- function(variable) {as.factor(ntile(variable, 5))}
```



# Retrieving and Wrangling Census, Transit, and Crime Data

#### Retrieve data from Census Bureau using tidycensus
```{r}
# Declare census variables of interest
census_var2000 <- c("P001001", "P006002", "PCT025009", "PCT025050", "P053001", "H056001", "P092001")
census_var2017 <- c("B25026_001E", "B02001_002E", "B15001_009E", "B15001_050E", "B19013_001E", "B25058_001E", "B06012_002E")
# total pop - p001001 or B25026_001E
# median age of males - p006002 (2000) or b02001_002e
# total male: 18-24 w/ bachelors degree - pct025009 or b15001_009e
# total female: 1824 w/ bachelors degree - pct025050 or b15001_050e
# median household income - p053001 or b19013_001e
# median contract rent - h056001 or b25058_001e
# total living in poverty - p092001 or b06012_002e

tractsDC.00 <- get_decennial(geography = "tract",
                                 year = 2000,
                                 variables = census_var2000,
                                 geometry = TRUE,
                                 state = 11999,
                                 output = "wide")

tractsDC.17 <- get_acs(geography = "tract", 
                           variables = census_var2017, 
                           year=2017, 
                           state=11999,
                           geometry=TRUE, 
                           output="wide")

```

#### Wrangle Census Data:
```{r}
#glimpse(tractsDC.00)

cleanTracts00 <- 
  tractsDC.00 %>%
  dplyr::select( -NAME, -geometry) %>%
  #spread(variable, estimate) %>%
  #st_drop_geometry() %>% #Drop geometry is the new way of removing the sf geometry class from the dataframe
	st_transform('ESRI:102746') %>%
  rename(TotalPop = P001001, 
  			 Whites = P006002, 
  			 MaleBachelors = PCT025009,
         FemaleBachelors = PCT025050, 
  			 MedHHInc = P053001, 
  			 MedRent = H056001,
         TotalPoverty = P092001) %>%
	mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2000") %>%
  dplyr::select(-Whites,-FemaleBachelors,-MaleBachelors,-TotalPoverty)

#glimpse(cleanTracts00)

#glimpse(tractsDC.17)

cleanTracts17 <- 
	tractsDC.17 %>%
	dplyr::select(-geometry, -B25026_001M, -B02001_002M, -B15001_009M, 
								-B15001_050M, -B19013_001M, -B25058_001M, -B06012_002M) %>% # Remove margin of error columns
	st_transform('ESRI:102746') %>%
  rename(TotalPop = B25026_001E, 
  			 Whites = B02001_002E,
         FemaleBachelors = B15001_050E, 
  			 MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E, 
  			 MedRent = B25058_001E,
         TotalPoverty = B06012_002E) %>%
  #dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2017") %>%
  dplyr::select(-Whites, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -NAME)

#glimpse(cleanTracts17)

allTracts <- rbind(cleanTracts00,cleanTracts17)
#glimpse(allTracts)

```

#### Retrieve & Clean Washington Metro Data
```{r}
# Although the Metro extends into multiple counties and cities in Virginia and Maryland, we will only examine stations in DC
metroStops <- st_read("https://opendata.arcgis.com/datasets/54018b7f06b943f2af278bbe415df1de_52.geojson")
metroLines <- st_read("https://opendata.arcgis.com/datasets/a29b9dbb2f00459db2b0c3c56faca297_106.geojson")

# We need to clean up the stops data a bit
# There are some stops in which multiple lines visit the stop. For our case, we will take the first stop listed
metroStops <- 
	metroStops %>%
	mutate(LINE = 
				 	plyr::revalue(LINE, 
						replace = c(
			 				"green, yellow" = "green", 
			 				"grn, yllw, orange, blue, slvr" = "green", 
				 			"red, green, yellow" = "red", 
			 				"red, blue, orange, silver" = "red",
				 			"blue, orange, silver" = "blue",
				 			"orange, blue, silver" = "orange"
					 	)
				 	)
				) %>%
	st_transform('ESRI:102746')

metroLines <-
	metroLines %>%
	st_transform('ESRI:102746')
	

```

# Map Metro Stops & Lines
```{r}

# Map of Metro Stops
ggplot() + 
  geom_sf(data=st_union(cleanTracts00)) +
  geom_sf(data=metroStops, 
          aes(colour = LINE), 
          show.legend = "point", size= 2) +
  scale_colour_manual(values = c("blue", "green", "orange", "red"), name = "Metro Line") +
  labs(title="Metro Stops", 
       subtitle="District of Columbia", 
       caption="Figure 1: Washington DC Metro Station locations, colored by Metro Line") +
  mapTheme()

# Map of Metro Lines
ggplot() +
	geom_sf(data = cleanTracts00) +
	geom_sf(data = metroLines,
					aes(color = NAME),
					show.legend = "line", size=2
					) +
	scale_color_manual(values = c("blue", "green", "orange", "red", "gray", "yellow"), name = "Metro Line") +
	labs(title = "Metro Lines",
			 subtitle = "District of Columbia",
			 ) +
	mapTheme()
```


# Analysis

#### Identifying TOD and Non-TOD Census Tracts:
```{r}
# First, lets define regions around each stop
# Create 0.5 mile (2640 foot) buffer around each transit station
# 0.5 mile is generally how far people are willing to walk to transportation
stopBuffers <- 
  rbind(
    st_buffer(metroStops, 2640) %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
    st_union(st_buffer(metroStops, 2640)) %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))

# Sometimes those regions overlap, we need to get rid of those overlaps
# Create an sf object with ONLY the unioned buffer
buffer <- filter(stopBuffers, Legend=="Unioned Buffer")

# Lets indentify the census tracts that overlap with the station buffered regions
# We will evaluate if the census tract centroid is within the station buffer region

### THIS IS AN INTERMEDIATE STEP AND MAY NOT NEED ITS OWN VARIABLE ###
selectCentroids <-
  st_centroid(cleanTracts00)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(cleanTracts00, GEOID)) %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")



# We want to be able to identify the tracts by whether they are TOD or Non-TOD
# Here we evaluate whether the census tract centroid is within the TOD buffer region
# and appropriately assign it a "TOD" or "Non-TOD" string
allTracts.group <- 
  rbind(
    st_centroid(allTracts)[buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD"))

# For visualization purposes down the line, we will create an sf object
# that has the outline of all TOD census tracts
#Create Outline of TOD Tracts for use in mapping
TOD_region <-
  allTracts.group%>%
    dplyr::select(TOD)%>%
  filter(TOD=="TOD") %>%
  st_union()%>%
      st_sf()

```


# Visualizizing Trends in Census Data:

#### TOD vs Non-TOD tracts across time
```{r}

# Lets visualize and compare the TOD/Non-TOD tracts across time
ggplot(allTracts.group)+
  geom_sf(data = st_union(cleanTracts17))+
  geom_sf(aes(fill = TOD)) +
  scale_fill_manual(values = c("grey80","red")) +
  labs(title = "TOD and Non-TOD Census Tracts",
       subtitle="Washington DC",
       caption="Figure 2: TOD and Non-TOD census tracts in 2000 and 2017. TOD census tracts are those where the 
                centroid is located within a half-mile of a Metro station") +
  facet_wrap(~year)+
  mapTheme() 

```

#### Population trends across space and time
```{r}

# Here we're just looking at Total Population with quantile breaks
ggplot() +
  geom_sf(data = cleanTracts00, aes(fill = q5(TotalPop))) +
  scale_fill_manual(values = palette5,
  									labels = qBr(cleanTracts00, "TotalPop"), 
                    name = "Popluation\n(Quintile Breaks)") +
  labs(title = "Total Population", subtitle = "District of Columbia; 2000") +
  mapTheme()

# Lets visualize and compare total population within the TOD/Non-TOD tracts across time
ggplot(allTracts.group)+
  geom_sf(data = st_union(cleanTracts17))+
	geom_sf(aes(fill = q5(TotalPop))) +
	geom_sf(data = TOD_region, fill="transparent", color ="red", size=0.5)+
  scale_fill_manual(values =  palette5,
  									labels = qBr(allTracts.group, "TotalPop"), 
                    name = "Popluation\n(Quintile Breaks)") +
  labs(title = "Total Population 2000-2017",
       subtitle="Washington DC",
       caption="Figure 3: Total population by census tract in 2000 and 2017") +
  facet_wrap(~year)+
  mapTheme() 

```